---
title: "BLUP"
author: "Peter von Rohr"
date: 2024-11-01
output:
  beamer_presentation:
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "asis")
knitr::knit_hooks$set(hook_convert_odg = rmdhelp::hook_convert_odg)
```

## Recap: Linear Mixed Effects Model

$$y = X \cdot b + Z \cdot u + e$$

\begin{tabular}{llp{8cm}}
where  &  &  \\
       &  $y$       &  vector of phenotypic observations of all animals \\
       &  $b$       &  vector of unknown fixed effects and unknown regression coefficients \\
       &  $u$       &  vector of random breeding values of all animals \\
       &  $e$       &  vector of random environmental effects \\
       &  $X$       &  design matrix with regression covariates or indicator values for fixed effects \\
       &  $Z$       &  design matrix relating breeding values to observations
\end{tabular}


## Solution - BLUP

* Solution to correction problem in selection index: BLUP
* Estimates environmental effects at the same time as breeding values are predicted
* Linear mixed effects model
* Meaning of BLUP
    + __B__ stands for __best__ $\rightarrow$ correlation between true ($u$) and its prediction ($\hat{u}$) is maximal or the prediction error variance ($var(u - \hat{u})$) is minimal.
    + __L__ stands for __linear__  $\rightarrow$ predicted breeding values are linear functions of the observations ($y$)
    + __U__ stands for __unbiased__ $\rightarrow$ expected values of the predicted breeding values are equal to the true breeding values
    + __P__ stands for __prediction__ 


## Example Dataset I

```{r, echo=FALSE, message=FALSE, warning=FALSE}
s_data_lm <- "https://charlotte-ngs.github.io/lbgfs2024/data/small_beef_data.csv"
tbl_lm <- readr::read_delim(s_data_lm, delim = ",")
knitr::kable(tbl_lm)
```

## Example Dataset II
```{r TableBeefExample, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
### # fix the numbers parents and offspring
n_nr_sire <- 3
n_nr_dam <- 8
n_nr_parents <- n_nr_sire + n_nr_dam
n_nr_offspring <- 16
n_nr_animals <- n_nr_parents + n_nr_offspring
### # assign parents to offspring and herds to records
vec_sire_id <- c(rep(1,8), rep(2,6), rep(3,2))
vec_dam_id <- rep(4:11,2)
vec_herd_codes <- c(rep(1,4), rep(2,4), rep(1,4), rep(2,4))
### # vector of observations
vec_weaning_weight <-  c(2.61,2.31,2.44,2.41,2.51,2.55,2.14,2.61,2.34,1.99,3.1,2.81,2.14,2.41,2.54,3.16)

### # create a tibble from the data
tbl_beef_data <- tibble::tibble( Animal = (n_nr_parents + 1):n_nr_animals,
                                    Sire   = vec_sire_id,
                                    Dam    = vec_dam_id[order(vec_dam_id)],
                                    Herd   = vec_herd_codes,
                                    `Weaning Weight` = vec_weaning_weight )
### # count number of observations
n_nr_observation <- nrow(tbl_beef_data)

### # parameters
h2 <- .25
n_var_p <- round(var(tbl_beef_data$`Weaning Weight`), digits = 4)
n_var_g <- round(h2 * n_var_p, digits = 4)
n_pop_mean <- round(mean(tbl_beef_data$`Weaning Weight`), digits = 2)

# save data to file, if needed
# s_data_path <- file.path(here::here(), "docs", "data", "beef_data_blup.csv")
# readr::write_delim(tbl_beef_data, file = s_data_path, delim = ",")

### # show the data frame
knitr::kable( tbl_beef_data, 
              booktabs = TRUE )
```


## Linear Models

* Simple linear model

$$y_{ij} = \mu + herd_j + e_{ij}$$

* Result: Estimate of effect of herd j
* Try with given dataset


## Linear Mixed Effects Model

* What about breeding value $u_i$ for animal $i$?
    + Problem: breeding values have a variance $\sigma_u^2$ and co-variances
    + Cannot be specified in simple linear model

$\rightarrow$ __Linear Mixed Effects Model__ (LME)

$$y_{ijk} = \mu + \beta_j + u_i + e_{ijk}$$


## Matrix-Vector Notation

* LME for all animals of a population

$\rightarrow$ use matrix-vector notation

$$y = X\beta + Zu + e$$

\begin{tabular}{llp{8cm}}
where  &  &  \\
       &  $y$      &  vector of length $n$ of all observations \\
       &  $\beta$  &  vector of length $p$ of all fixed effects  \\
       &  $X$      &  $n \times p$ design matrix linking the fixed effects to the observations \\
       &  $u$      &  vector of length $n_u$ of random effects \\
       &  $Z$      &  $n \times n_u$ design matrix linking random effect to the observations \\
       &  $e$      &  vector of length $n$ of random residual effects.  
\end{tabular}


## Expected Values and Variances

* Expected values

$$E(u) = 0 \text{  and  } E(e) = 0 \rightarrow E(y) = X\beta$$

* Variances

$$var(u) = U  \text{  and  } var(e) = R$$

with $cov(u, e^T) = 0$, $$var(y) = Z * var(u) * Z^T + var(e) = ZUZ^T + R = V$$


## Estimates of unknown Parameters

Use properties of BLUP to design predictor $\widehat{u}$

* Linear: predictor $\widehat{u}$ is a linear function of data

$$\widehat{u} = M \cdot (y - k)$$

* Unbiased: $E(\widehat{u}) = E(u)$. Above we have specified $E(u) = 0$

$$k = E(y)$$

* Best: Prediction error variance $var(u - \widehat{u})$ is minimal

$$M = UZ^TV^{-1}$$

## Result

* Summary of above results, using $E(y) = X\beta$

$$\hat{u} = UZ^TV^{-1}(y - X\beta)$$

* For $u$ and $y$ following a Gaussian normal distribution

$$\hat{u} = E(u|y)$$

* $\beta$ unknown, replace it by general least squares estimate 

$$\hat{\beta} = (X^T V^{-1} X)^- X^T V^{-1} y$$


## Mixed Model Equations

* Problem: $V^{-1}$
* Same solutions obtained with following set of equations

$$\left[
  \begin{array}{lr}
  X^T R^{-1} X  &  X^T R^{-1} Z \\
  Z^T R^{-1} X  &  Z^T R^{-1} Z + U^{-1}
  \end{array}
\right]
\left[
  \begin{array}{c}
  \hat{\beta} \\
  \hat{u}
  \end{array}
\right]
=
\left[
  \begin{array}{c}
  X^T R^{-1} y \\
  Z^T R^{-1} y
  \end{array}
\right]$$


## Sire Model

* Breeding value of sire as random effect: 

$$y = X\beta + Zs + e$$
* Expected values

$$E(s) = 0 \text{  and  } E(e) = 0 \rightarrow E(y) = X\beta$$

* Variances

$$var(s) = U_s  \text{  and  } var(e) = R$$

* Unrelated sires: $var(s) = U_s = I \sigma_s^2$

## Example
```{r siremodelbeefexample, echo=FALSE, results='asis'}
mat_x_sire <- matrix(data = c(1, 0,
                              1, 0,
                              1, 0,
                              1, 0,
                              0, 1,
                              0, 1,
                              0, 1,
                              0, 1,
                              1, 0,
                              1, 0,
                              1, 0,
                              1, 0,
                              0, 1,
                              0, 1,
                              0, 1,
                              0, 1), ncol = 2, byrow = TRUE)
vec_betahat_sire <- c("\\beta_1", "\\beta_2")
mat_z_sire <- matrix(data = c(1, 0, 0,
                              1, 0, 0,
                              1, 0, 0,
                              1, 0, 0,
                              1, 0, 0,
                              1, 0, 0,
                              1, 0, 0,
                              1, 0, 0,
                              0, 1, 0,
                              0, 1, 0,
                              0, 1, 0,
                              0, 1, 0,
                              0, 1, 0,
                              0, 1, 0,
                              0, 0, 1,
                              0, 0, 1), ncol = 3, byrow = TRUE)
vec_sirehat_sire <- c("s_1", "s_2", "s_3")
vec_res_sire <- c("e_1", "e_2", "e_3", "e_4", "e_5", "e_6", "e_7", "e_8", "e_9", "e_{10}", "e_{11}", "e_{12}", "e_{13}", "e_{14}", "e_{15}", "e_{16}")
cat("$$ \n")
# cat(paste(rmdhelp::bcolumn_vector(pvec_avector = vec_weaning_weight), sep = "\n"),"\n")
cat(paste(rmdhelp::bcolumn_vector(pvec = vec_weaning_weight), sep = "\n"), "\n")
cat("= \n")
cat(paste(rmdhelp::bmatrix(pmat = mat_x_sire), sep = "\n"), "\n")
cat(paste(rmdhelp::bcolumn_vector(pvec = vec_betahat_sire), sep = "\n"), "\n")
cat("+ \n")
cat(paste(rmdhelp::bmatrix(pmat = mat_z_sire), sep = "\n"), "\n")
cat(paste(rmdhelp::bcolumn_vector(pvec = vec_sirehat_sire), sep = "\n"), "\n")
cat("+ \n")
cat(paste(rmdhelp::bcolumn_vector(pvec = vec_res_sire), sep = "\n"), "\n")
cat("$$ \n")
```


## Animal Model

* Breeding value for all animals as random effects

$$y = X\beta + Zu + e$$

* Expected values

$$E(u) = 0 \text{  and  } E(e) = 0 \rightarrow E(y) = X\beta$$

* Variances

$$var(u) = U = A * \sigma_u^2  \text{  and  } var(e) = R$$

* Matrix $A$: numerator relationship matrix and $\sigma_u^2$ the genetic additive variance