---
title: "LBGFS2024 - Problem 3: Inbreeding"
output: html_notebook
---


```{r setup, include=FALSE}
params <- list(seed = 2417)
```

## Required Package `pedsuite`

```{r}
if (!is.element("pedsuite", installed.packages()))
  install.packages("pedsuite", dependencies = TRUE)
```


## Random Pedigree 
Start with a random pedigree

```{r}
ped_p3 <- pedtools::randomPed(8, founders = 3, seed = params$seed)
ped_p3
```

Show the plot

```{r}
plot(ped_p3)
```


Add more animals, start wit a son

```{r}
ped_p3 <- ped_p3 |>
  addSon(parents = c(7,8))
plot(ped_p3)
```

Add a daughter

```{r}
ped_p3 <- ped_p3 |>
  addDaughter(parents = c(9,6))
plot(ped_p3)
```


## Convert To tibble

```{r}
tbl_ped_p3 <- tibble::as.tibble(ped_p3)
tbl_ped_p3
```

## Convert To pedigreemm

```{r}
# replace MVC from 0 to NA
tbl_ped_p3$fid[tbl_ped_p3$fid == 0] <- NA
tbl_ped_p3$mid[tbl_ped_p3$mid == 0] <- NA
# convert to pedigreemm
pem_p3 <- pedigreemm::pedigree(sire = as.integer(tbl_ped_p3$fid),
                               dam = as.integer(tbl_ped_p3$mid),
                               label = tbl_ped_p3$id)
pem_p3
```

## Inbreeding

```{r}
pedigreemm::inbreeding(ped = pem_p3)
```


## Write Pedigree to Data File
Filter to write only rows with non-founders

```{r}
library(dplyr)
tbl_ped_p3_data <- tbl_ped_p3 %>%
  filter(!is.na(fid)) %>%
  filter(!is.na(mid)) %>%
  rename(Animal = id) %>%
  rename(Sire = fid) %>%
  rename(Dam = mid) %>%
  rename(Sex = sex) %>%
  select(Animal, Sire, Dam)
tbl_ped_p3_data
```


```{r}
s_ped_p3_data <- here::here("docs", "data", "exam_pedigree_p3.csv")
readr::write_delim(tbl_ped_p3_data, s_ped_p3_data, delim = ",")
```


## Restart Analysis from Data Read from File

```{r, message=FALSE, warning=FALSE}
rm(list = ls())
s_data_root <- "https://charlotte-ngs.github.io/lbgfs2024/data"
s_ped_p3 <- file.path(s_data_root, "exam_pedigree_p3.csv")
tbl_ped_p3 <- readr::read_delim(s_ped_p3, delim = ",")
tbl_ped_p3
```


### Augment Pedigree
The pedigree augmentation consist of two steps, first find founders and then add records using the following two functions

```{r}
# function to get founder animals
get_founders <- function(ptbl_ped){
  # sire founders
  vec_sire_fnd <- setdiff(ptbl_ped$Sire, ptbl_ped$Animal)
  # dam founders
  vec_dam_fnd <- setdiff(ptbl_ped$Dam, ptbl_ped$Animal)
  # combine
  vec_founders <- c(vec_sire_fnd, vec_dam_fnd)
  # remove NA
  vec_founders <- vec_founders[!is.na(vec_founders)]
  # order
  vec_founders <- vec_founders[order(vec_founders)]
  # return
  return(vec_founders)
}
# add founders to pedigree
augment_pedigree <- function(ptbl_ped){
  # get founders
  vec_founders <- get_founders(ptbl_ped = ptbl_ped)
  n_nr_founders <- length(vec_founders)
  # add founder records
  tbl_aug_ped <- dplyr::bind_rows(
    tibble::tibble(Animal = vec_founders,
                   Sire = rep(NA, n_nr_founders),
                   Dam = rep(NA, n_nr_founders)),
    ptbl_ped)
  # return result
  return(tbl_aug_ped)
}
```

The autmented form

```{r}
tbl_ped_p3_aug <- augment_pedigree(ptbl_ped = tbl_ped_p3)
tbl_ped_p3_aug
```

### Pedigreemm-Format

```{r}
pem_p3 <- pedigreemm::pedigree(sire = tbl_ped_p3_aug$Sire,
                              dam = tbl_ped_p3_aug$Dam,
                              label = tbl_ped_p3_aug$Animal)
pem_p3
```

Inbreeding coefficients are computed as 

```{r}
(vec_inb <- pedigreemm::inbreeding(ped = pem_p3))
```

The table for the task

```{r}
vec_status <- sapply(vec_inb, function(x) ifelse(x > 0, "Y", "N"), USE.NAMES = F)
tbl_inb_sol <- tibble::tibble(Animal = tbl_ped_p3_aug$Animal,
                              Inbred = vec_status,
                             `Inbreeding Coefficient` = vec_inb)
tbl_inb_sol
```


## Problem 3b: Min Inbreeding
For every sire, find the dam such that the inbreeding coefficient of the resulting offspring is minimal

This is easiest done by looking at the numerator relationship matrix

```{r}
mat_A <- as.matrix(pedigreemm::getA(ped = pem_p3))
mat_A
```

* vector of sires and dams

```{r}
vec_sire <- unique(tbl_ped_p3$Sire)
(vec_sire <- vec_sire[order(vec_sire)])
vec_dam <- unique(tbl_ped_p3$Dam)
(vec_dam <- vec_dam[order(vec_dam)])
```

Check for each sire which dam has minimal relationship coef

```{r}
n_inb_limit <- 0.25
for (sidx in vec_sire){
  vec_dam_mates <- which(mat_A[sidx,vec_dam]/2 < n_inb_limit)
  if (length(vec_dam_mates) > 0){
    cat (" * Mates for sire ", sidx, ": ", paste0(vec_dam[vec_dam_mates], collapse = ", "), "\n")
  } else {
    cat (" * No mates found for sire: ", sidx, "\n")
  }
}
```

