---
title: "Prepare Genotypic Value Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prepare Genotypic Value Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Disclaimer
Dataset to estimate genotypic values from data is prepared. This dataset will be used in lbg_ex02.


# Genotype Data
Genotypes are taken from the dataset of lbg_ex01 described in `prepare_geno_data.Rmd`.

```{r}
s_csv_data_path <- here::here("docs", "data", "geno_first_100.csv")
tbl_geno <- readr::read_delim(s_csv_data_path, delim = ",")
tbl_geno
```

The genotype data is reduced to only contain the columns with the IDs and the column of Locus 3, as this was found to be the first locus having all genotypes.

```{r}
library(dplyr)
tbl_geno_val <- tbl_geno %>% select(Ids,Loc_3)
tbl_geno_val
```


# Simulation of Values
Initialise values to be only residuals

```{r}
set.seed(1004)
n_mean_res <- 0
n_sd_res <- 1.5
tbl_geno_val$y <- rnorm(nrow(tbl_geno_val), mean = n_mean_res, sd = n_sd_res)
tbl_geno_val
```

Generate genotypic values

```{r}
# vector of values
n_par_a <- 13.3
n_par_d <- 4.9
vec_geno_val <- c(-n_par_a, n_par_d, n_par_a)
# genotype matrix
mat_geno_X <- model.matrix(y ~ 0 + as.factor(Loc_3), data = tbl_geno_val)
attr(mat_geno_X, "assign") <- NULL
attr(mat_geno_X, "contrasts") <- NULL
colnames(mat_geno_X) <- NULL
mat_geno_val <- crossprod(t(mat_geno_X), vec_geno_val)
mat_geno_val
```

Add genotypic values to residuals

```{r}
tbl_geno_val$y <- tbl_geno_val$y + mat_geno_val[,1]
tbl_geno_val
```


# Plot
Plot values against genotypes

```{r}
plot(tbl_geno_val$Loc_3, tbl_geno_val$y)
```

The same plot with `ggplot2`

```{r}
library(ggplot2)
p_geno <- ggplot(data = tbl_geno_val, aes(x = Loc_3, y = y)) + 
  geom_point()
p_geno
```


# Writing Data To File
Rename column with genotypes from `Loc_3` to `Loc_G`.

```{r}
tbl_geno_val <- tbl_geno_val %>% rename(Loc_G = Loc_3)
tbl_geno_val
```

Round observation values

```{r}
tbl_geno_val$y <- round(tbl_geno_val$y, digits = 4)
```


Write results to file

```{r}
s_geno_val_path <- here::here("docs", "data", "geno_val_single_loc.csv")
readr::write_delim(tbl_geno_val, s_geno_val_path, delim = ",")
```


