---
title: "LBG - FS2024 -- Exercise 11"
author: Peter von Rohr
format: live-html
engine: knitr
webr:
  packages:
    - readr
---

{{< include ../../_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
#| message: false
#| warning: false 
s_data_url_path <- "https://charlotte-ngs.github.io/lbgfs2024/data/geno_data.csv"
tbl_all_data <- readr::read_delim(s_data_url_path, delim = ",")
n_nr_animal <- nrow(tbl_all_data)
n_nr_snp <- length(grep("SNP", colnames(tbl_all_data), fixed = T))
lambda_g <- 3
```


## Problem 1: Marker Effect Model
We are given the dataset that is shown in the table below. This dataset contains gentyping results of `r n_nr_animal` animals for `r n_nr_snp` SNP loci.

```{r}
#| echo: false
#| message: false
#| warning: false 
#| results: asis
cat("```{webr}\n")
cat("#| message: false\n")
cat("#| warning: false\n") 
cat("#| edit: false\n")
cat(paste0("s_data_url_path <- '", s_data_url_path, "'\n"))
cat("tbl_all_data <- readr::read_delim(s_data_url_path, delim = ',')\n")
cat("tbl_all_data\n")
cat("```\n")
```



### Task
* The goal of this problem is to estimate SNP marker effects using a `marker effect model`. Because we have just `r n_nr_snp` SNP loci, you can use a fixed effects linear model with the `r n_nr_snp` loci as fixed effects. Furthermore you can also include a fixed intercept into the model.
* You can use the R-function `lm()` to get the solutions for estimates of the unknown SNP effects.
* Compute genomic breeding values for all animals with genotypic data


### Solution 

* Use a regression model for the SNP-effects

```{webr}
# fit regression model to get marker effects

```


* Use marker effects to compute genomic breeding values, start by extracting estimates of marker effects

```{webr}
# extract marker effects from summary of lm

```

* Compute direct genomic breeding values

```{webr}
# extract matrix with SNP-genotypes from dataset

# compute genomic breeding values

```

* Rank animals according to genomic breeding values

```{webr}
# order animals according to genomic breeding values

```



## Problem 2: Breeding Value Model
Use the same data as in Problem 1 to estimate genomic breeding values using a `breeding value model`.


### Hints
* The only fixed effect in this model is the mean $\mu$ which is the same for all observations.
* You can use the following function to compute the inverse genomic relationship matrix

```{webr}
#| edit: false
#' Compute genomic relationship matrix based on data matrix
computeMatGrmInv <- function(pmatData) {
  matData <- pmatData
  # check the coding, if matData is -1, 0, 1 coded, then add 1 to get to 0, 1, 2 coding
  if (min(matData) < 0) matData <- matData + 1
  # Allele frequencies, column vector of P and sum of frequency products
  freq <- apply(matData, 2, mean) / 2
  P <- 2 * (freq - 0.5)
  sumpq <- sum(freq*(1-freq))
  # Changing the coding from (0,1,2) to (-1,0,1) and subtract matrix P
  Z <- matData - 1 - matrix(P, nrow = nrow(matData), 
                             ncol = ncol(matData), 
                             byrow = TRUE)
  # Z%*%Zt is replaced by tcrossprod(Z)
  matG <- tcrossprod(Z)/(2*sumpq)
  matG_result <- matG + 0.05 * diag(nrow = nrow(matG))
  # more regularization, if needed
  n_max_count <- 10
  n_cur_count <- 0
  n_lim_min_eig <- .Machine$double.eps^0.25
  while (min(eigen(matG_result, only.values = T)$values) < n_lim_min_eig && n_cur_count < n_max_count){
    matG_result <- matG_result + 0.025 * diag(nrow = nrow(matG))
    n_cur_count <- n_cur_count + 1
  }
  if (min(eigen(matG_result, only.values = T)$values) < n_lim_min_eig) 
    stop(" *** ERROR: CANNOT FIND invertible matrix, min. eigen: ", min(eigen(matG_result, only.values = T)$values))
  return(solve(matG_result))
}
```


### Solution

* Design matrix $X$ for fixed effects

```{webr}
# design matrix X

```

* Design matrix $W$ for random genomic breeding values

```{webr}
# design matrix W

```

* Get genomic relationship matrix

```{webr}
# compute inverse genomic relationship matrix

```

* Set up coefficient matrix assuming a value of $\lambda = \sigma_e^2 / \sigma_u^2 = `r lambda_g`$

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: asis
cat("```{webr}\n")
cat("# initialise lambda\n")
cat(paste0("lambda_g <- ", lambda_g, "\n"))
cat("# set up coefficient matrix of mme\n")
cat("\n")
cat("# set up right-hand side\n")
cat("\n")
cat("```\n")
```

* Get list of genomic breeding values

```{webr}
# select solutions for genomic breeding values

```

* Ranking animals according to their genomic breeding values

```{webr}
# rank animals

```


