---
title: "LBGFS2024 - Problem 2: Numerator Relationship Matrix"
output: html_notebook
---

```{r setup, include=FALSE}
params <- list(seed = 2417)
```

## Required Package `pedsuite`

```{r}
if (!is.element("pedsuite", installed.packages()))
  install.packages("pedsuite", dependencies = TRUE)
```


## Create A complex family tree as a test
Start with a half-sib pedirgee

```{r}
library(pedtools)
ped_p2a <- halfSibPed(nch1 = 2, nch2 = 3, sex1 = 2:1, sex2 = c(2, 1, 1)) |> 
  addDaughter(parents = 5:6) |>
  addSon(parents = c(4,7)) |>
  addDaughter(parents = c(7,9)) |>
  addDaughter(parents = c(8,11))
plot(ped_p2a)
```

Convert the pedigree to a tibble

```{r}
tbl_ped_p2a <- tibble::as.tibble(ped_p2a)
tbl_ped_p2a
```

Rename columns to 

```{r}
tbl_ped_p2a <- dplyr::rename(tbl_ped_p2a, Animal = id, Sire = fid, Dam = mid)
tbl_ped_p2a
```

Remove founder animals

```{r}
library(dplyr)
tbl_ped_p2a_data <- tbl_ped_p2a %>% 
  filter(Sire != "0") %>%
  filter(Dam != "0") %>% 
  select(Animal, Sire, Dam)
tbl_ped_p2a_data
```

## Write Data To File

```{r}
s_ped_p2_data <- here::here("docs", "data", "exam_pedigree_p2.csv")
readr::write_delim(tbl_ped_p2a_data, s_ped_p2_data, delim = ",")
```


## Random Pedigree 
For problem 2b, we use a random pedigree. 

```{r}
ped_p2b <- pedtools::randomPed(11, founders = 4, seed = params$seed)
ped_p2b
```

Show the graphical presentation

```{r}
plot(ped_p2b)
```

Change unknown parents from "*" to NA 

```{r}
# change "*" for unknown parents
tbl_ped_p2b <- tibble::as.tibble(ped_p2b)
tbl_ped_p2b$fid[tbl_ped_p2b$fid == "0"] <- NA
tbl_ped_p2b$mid[tbl_ped_p2b$mid == "0"] <- NA
tbl_ped_p2b
```

Get inverse of $A$

```{r}
pem_p2b <- pedigreemm::pedigree(sire = as.integer(tbl_ped_p2b$fid),
                               dam = as.integer(tbl_ped_p2b$mid),
                               label = tbl_ped_p2b$id)
mat_A_inv <- pedigreemm::getAInv(pem_p2b)
mat_A_inv
```


## Write Matrix To File

```{r}
tbl_mat_A_inv <- tibble::as.tibble(as.matrix(mat_A_inv))
s_mat_p2 <- here::here("docs", "data", "exam_inv_num_rel_mat_p2.csv")
readr::write_delim(tbl_mat_A_inv, s_mat_p2, delim = ",")
```


## Matrix Decomposition of $A$
The cholesky decomposition of $A$ is given by

$$A = R\cdot R^T$$
where $R$ is a lower diagonal matrix. This can be obtained with

```{r}
mat_A <- as.matrix(pedigreemm::getA(pem_p2b))
mat_R <- t(chol(mat_A))
mat_R
```

Check

```{r}
round(mat_R %*% t(mat_R) - mat_A, digits = 8)
```


The matrix $R$ can be decomposed into 

$$R = L\cdot S$$
where $L$ is a lower triangular matrix with a diagonal of all ones and $S$ is the diagonal matrix with diagonal elements which are the same as the matrix $R$. Hence $S$ is obtained by 

```{r}
mat_S <- diag(diag(mat_R), nrow = nrow(mat_R))
mat_S
```

Matrix $L$ can be computed as 

$$L = R \cdot S^{-1}$$

```{r}
mat_L = mat_R %*% solve(mat_S)
mat_L
```

The matrix $P$ is obtained as $P = I - L^{-1}$

```{r}
mat_P <- diag(nrow = nrow(mat_L)) - solve(mat_L)
round(mat_P, digits = 6)
```


### Port to $A^{-1}$
Solution via decomposition of $A^{-1}$ does not work. The following sections show why. 

Based on the cholesky decomposition of $A = R \cdot R^T$, we get 

$$A^{-1} = (R^T)^{-1} \cdot R^{-1}$$

which is a valid decomposition, but it is not a cholesky decomposition, because $(R^T)^{-1}$ is not a lower triangular matrix. 

The cholesky decomposition of $A^{-1}$ corresponds to

$$A^{-1} = Q \cdot Q^T$$
where $Q$ is lower triangular. But this matrix cannot be set equal to $(R^T)^{-1}

This leads to the relationship between matrices $Q$ and $R$ as 

$$Q \ne (R^T)^{-1} = (R^{-1})^T$$



```{r}
mat_A_inv <- as.matrix(pedigreemm::getAInv(pem_p2b))
mat_A_inv
```

The cholesky decomposition factor $Q$ of $A^{-1}$ is then

```{r}
mat_Q <- t(chol(mat_A_inv))
mat_Q
```

which is clearly lower triangular. The following shows that $(R^T)^{-1}$ is upper triangular and hence cannot be equal to $Q$. 

```{r}
solve(t(mat_R))
```


Check

```{r}
round(mat_Q %*% t(mat_Q) - mat_A_inv, digits = 8)
```

```{r}
round(solve(t(mat_R)) %*% solve(mat_R) - mat_A_inv, digits = 8)
```



```{r}
round(mat_Q - solve(t(mat_R)), digits = 8)
```

```{r}
round(mat_Q - solve(mat_R), digits = 8)
```



Decomposition of $Q$ into 

```{r}
mat_S_inv <- diag(diag(mat_Q), nrow = nrow(mat_Q))
mat_S_inv
```

Getting $(L^T)^{-1} = Q \cdot S$ as 

```{r}
mat_L_inv_t <- mat_Q %*% solve(mat_S_inv)
mat_L_inv_t
```


The cholesky decomposition is given by 

$$A^{-1} = RR^T$$

A triangular matrix $R$ can be decomposed into a diagonal matrix $S$ and a triangular matrix $L$. 

$$R = LS$$

The matrix $S$ is the diagonal elements of $R^T$, then $L$ is computed as 

$$L = RS^{-1}$$


$$A^{-1} = LDL^T$$

```{r}
mat_A_inv <- as.matrix(mat_A_inv)
mat_A_inv
```

The cholesky decomposition of $A^{-1}$ is 


```{r}
mat_R <- t(chol(mat_A_inv))
mat_R
```


The matrix $S$ is 

```{r}
mat_S <- diag(diag(mat_R), nrow = nrow(mat_R))
mat_S
```

The matrix $L^{-1}$ is computed as 

```{r}
mat_L <- mat_R %*% solve(mat_S)
mat_L
```

```{r}
mat_P <- diag(nrow = nrow(mat_A_inv)) - solve(mat_L)
mat_P
```

Check whether it works with matrix $A = L \cdot D \cdot L^T$

```{r}
mat_A <- as.matrix(pedigreemm::getA(ped = pem_p2b))
mat_A
```

The cholesky decomposition of $A$ is: $A = RR^T$

```{r}
mat_R <- t(chol(mat_A))
mat_R
```

The decomposition of $R = L\cdot S$

```{r}
mat_S <- diag(diag(mat_R), nrow = nrow(mat_A))
mat_S
```

The $L$ is given by $L = R \cdot S^{-1}$

```{r}
mat_L <- mat_R %*% solve(mat_S)
mat_L
```

Check

```{r}
mat_D <- mat_S %*% mat_S
mat_A_comp <- mat_L %*% mat_D %*% t(mat_L)
mat_A_comp
```

```{r}
mat_A - mat_A_comp
```

```{r}
mat_P = diag(nrow = nrow(mat_A)) - solve(mat_L)
round(mat_P, digits = 6)
```

Check

```{r}
mat_L_inv <- solve(mat_L)
mat_L_inv
```

Multiply back

```{r}
mat_D_inv <- solve(mat_D)
mat_D_inv
```

```{r}
1/pedigreemm::Dmat(ped = pem_p2b)
```


```{r}
mat_A_inv_comp <- t(mat_L_inv) %*% mat_D_inv %*% mat_L_inv
mat_A_inv_comp
```

Compare

```{r}
round(mat_A_inv - mat_A_inv_comp, digits = 6)
```


