---
title: "Livestock Breeding and Genomics"
---

```{r}
#| label: versionextractor
#| echo: false
# constants
s_quarto_yml_file <- "_quarto.yml"
s_title_sep <- " | "
# extract version from DESC
s_cur_version <- desc::desc_get_version()
l_quarto_yml <- yaml::read_yaml(s_quarto_yml_file)
s_title <- l_quarto_yml$website$title
vec_title_comp <- unlist(strsplit(s_title, split = s_title_sep, fixed = T))
l_quarto_yml$website$title <- paste0(vec_title_comp[1], s_title_sep, s_cur_version, collapse = "")
# fix problem with toc
l_quarto_yml$format$html$toc <- 'true'
yaml::write_yaml(l_quarto_yml, file = s_quarto_yml_file)
# remove quote
vec_yml <- readLines(file(s_quarto_yml_file))
writeLines(gsub("'", "", vec_yml, fixed = T), file(s_quarto_yml_file))
# news file
s_news_md <- here::here("News.md")
s_news_qmd <- "weekly_news.qmd"
vec_news_md <- readLines(con = file(s_news_md))
vec_header_lines <- grep("^#", vec_news_md)
if (length(vec_header_lines) > 0){
  # loop over header lines and add week numbering
  s_prev_week <- ""
  for (l in vec_header_lines){
    s_version <- unlist(strsplit(vec_news_md[l], " "))[3]
    s_current_week <- unlist(strsplit(s_version, ".", fixed = T))[2]
    if (s_prev_week != s_current_week){
      vec_news_md[l] <- paste0("# Week ", s_current_week, " \n", "#", vec_news_md[l], collapse = "")
      s_prev_week <- s_current_week
    } else {
      vec_news_md[l] <- paste0("#", vec_news_md[l], collapse = "")
    }
  }
} else {
  cat(" *** CANNOT FIND header lines in: ", s_news_md)
}
writeLines(vec_news_md, con = file(s_news_qmd))
# add footer
if (is.element("rmdhelp", installed.packages()))
  cat(paste0(rmdhelp::get_latest_change()), file = s_news_qmd, append = T)

```


## Course Catalogue
The link to the course catalogue: [751-6305-00L  Livestock Breeding and Genomics](https://www.vvz.ethz.ch/Vorlesungsverzeichnis/lerneinheit.view?lerneinheitId=182596&semkez=2024W&ansicht=LEHRVERANSTALTUNGEN&lang=en)


## Topics for Master Theses
Topics for Master Theses at Qualitas AG together with ETH are available [here](https://charlotte-ngs.github.io/lbgfs2024/topics_msc_intern/topics_msc_intern.html)


## Information
- Lecturer: Peter von Rohr
- Date: Friday 9-12
- Room: LFW C11 
- Moodle: https://moodle-app2.let.ethz.ch/course/view.php?id=23928
- Zoom: https://ethz.zoom.us/j/66360111190


### Exercise Platform
Three options are available

1. Webr: [https://webr.r-wasm.org/latest](https://webr.r-wasm.org/latest)
2. Posit cloud: [https://posit.cloud](https://posit.cloud)
3. Run your own platform: Follow steps on [https://posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop)


### Exam
- Written exam
    + Date: 20.12.2024 - 09:15 - 11:15
    + Room: LFW C 11
    + [Datasets](https://charlotte-ngs.github.io/lbgfs2024/exam/20241220_exam_datasets.html) for written exam
- Aural exam
    + Date: 17.01.2025 - 10:00 - 11:00
    + Exam Problems


## Questions
Questions can be asked during the lecture and during the exercise hour or via e-mail: 

- Peter von Rohr (peter.vonrohr `at` usys.ethz.ch)


## Lecture Program
```{r}
#| echo: false
### # header names of tables
vecTableHeaders <- c("Week", "Date", "Topic")
# define course start date
dCourseStart <- as.Date("2024/09/20")
# set number of weeks in semester
nNrSemesterWeeks <- 14
# define columns for weaks, dates, and subjects
Week <- 1:nNrSemesterWeeks
Date <- format(seq(dCourseStart, by = "week", length.out = nNrSemesterWeeks), "%d.%m")
Topic <- vector(mode = "character", length = nNrSemesterWeeks)
# subjects per week
Topic[1] <- "Introduction to Livestock Breeding and Genomics"
Topic[2] <- "Review of Quantitative Genetics/Single Locus"
Topic[3] <- "Genetic Evaluation with Different Sources of Information"
Topic[4] <- "Genetic Covariance Between Relatives"
Topic[5] <- "Best Linear Unbiased Prediction (BLUP) - Univariate Analysis"
Topic[6] <- "BLUP - Additional Aspects"
Topic[7] <- "BLUP - Multiple Traits"
Topic[8] <- "Variance and Inbreeding"
Topic[9] <- "Variance Components Estimation"
Topic[10] <- "Genomic Selection"
Topic[11] <- "Genom-Wide Association Studies"
Topic[12] <- "Review on Selection Index Theory"
Topic[13] <- "Test-Exam and Questions"
Topic[14] <- "Exam"

dfCnTable <- data.frame(Week, Date, Topic, stringsAsFactors = FALSE)
colnames(dfCnTable) <- vecTableHeaders
knitr::kable(dfCnTable)
```



```{r}
#| echo: false
#| results: asis
if (is.element("rmdhelp", installed.packages()))
  cat(paste0(rmdhelp::get_latest_change()))
```

