---
title: ""
format: html
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE)
knitr::knit_hooks$set(hook_convert_odg = rmdhelp::hook_convert_odg)
```

```{bash}
#| echo: false
#| eval: false
# bash - export as jpeg
ls -1 "/Users/${USER}/ngs.charlotte@gmail.com - Google Drive/My Drive/Vorlesung/LBG/FS2024/OHP-L01"/* | while read p
do
  CURPICFN=$(basename "$p")
  echo " * Copy $p ..."
  cp "$p" ~/Desktop
  echo " * Convert ~/Desktop/$CURPICFN ..."
  mogrify -resize 50% -format jpg ~/Desktop/$CURPICFN
  echo " * Remove ~/Desktop/$CURPICFN ..."
  rm ~/Desktop/$CURPICFN
  sleep 2
done
```

```{r}
#| echo: false
#| eval: false
# include jpg into this rmd
this_rmd_file <- ifelse(rstudioapi::isAvailable(),
                        rstudioapi::getActiveDocumentContext()$path,
                        whereami::thisfile())


vec_pics <- list.files(path = "~/Desktop", pattern = ".jpg$")
#vec_pics <- vec_pics[c(1:6, 10:11, 7:9)]
for (pidx in seq_along(vec_pics)){
   cur_pic <- vec_pics[pidx]
   cur_pic_nr <- as.integer(unlist(strsplit(fs::path_ext_remove(cur_pic), "_"))[2])
  cat("\n## OHP Picture ", pidx, "\n\n", sep = "", file = this_rmd_file, append = TRUE)
  cat("```{r ohp-picture", cur_pic_nr, "}\n", sep = "", file = this_rmd_file, append = TRUE)
  # cat("```{r}\n", sep = "", file = this_rmd_file, append = TRUE)
  # cat("#| label: ohp-picture", cur_pic_nr, "\n", sep = "", file = this_rmd_file, append = TRUE)
  cat("rmdhelp::use_odg_graphic(ps_path = 'odg/ohp-page", cur_pic_nr, ".odg', ps_odg_template = 'odg_draw_a4_portrait')\n", sep = "", file = this_rmd_file, append = TRUE)
  cat("if (fs::file_exists('~/Desktop/", cur_pic, "')) fs::file_delete(path = '~/Desktop/", cur_pic, "')\n", sep = "", file = this_rmd_file, append = TRUE)
  cat("```\n\n", file = this_rmd_file, append = TRUE)
}
```


## OHP Picture 1

```{r ohp-picture4302}
rmdhelp::use_odg_graphic(ps_path = 'odg/ohp-page4302.odg', ps_odg_template = 'odg_draw_a4_portrait')
if (fs::file_exists('~/Desktop/IMG_4302.jpg')) fs::file_delete(path = '~/Desktop/IMG_4302.jpg')
```


## OHP Picture 2

```{r ohp-picture4303}
rmdhelp::use_odg_graphic(ps_path = 'odg/ohp-page4303.odg', ps_odg_template = 'odg_draw_a4_portrait')
if (fs::file_exists('~/Desktop/IMG_4303.jpg')) fs::file_delete(path = '~/Desktop/IMG_4303.jpg')
```


## OHP Picture 3

```{r ohp-picture4304}
rmdhelp::use_odg_graphic(ps_path = 'odg/ohp-page4304.odg', ps_odg_template = 'odg_draw_a4_portrait')
if (fs::file_exists('~/Desktop/IMG_4304.jpg')) fs::file_delete(path = '~/Desktop/IMG_4304.jpg')
```


## OHP Picture 4

```{r ohp-picture4305}
rmdhelp::use_odg_graphic(ps_path = 'odg/ohp-page4305.odg', ps_odg_template = 'odg_draw_a4_portrait')
if (fs::file_exists('~/Desktop/IMG_4305.jpg')) fs::file_delete(path = '~/Desktop/IMG_4305.jpg')
```


## OHP Picture 5

```{r ohp-picture4306}
rmdhelp::use_odg_graphic(ps_path = 'odg/ohp-page4306.odg', ps_odg_template = 'odg_draw_a4_portrait')
if (fs::file_exists('~/Desktop/IMG_4306.jpg')) fs::file_delete(path = '~/Desktop/IMG_4306.jpg')
```


## OHP Picture 6

```{r ohp-picture4307}
rmdhelp::use_odg_graphic(ps_path = 'odg/ohp-page4307.odg', ps_odg_template = 'odg_draw_a4_portrait')
if (fs::file_exists('~/Desktop/IMG_4307.jpg')) fs::file_delete(path = '~/Desktop/IMG_4307.jpg')
```

