---
title: "Derive Hendersons Rules for Setting Up $A^{-1}$"
output: html_notebook
---


# Pedigree
The pedigree is grown gradually. We start with two unrelated animals $1$ and $2$. This means

```{r}
n_nr_founder <- 2
tbl_ped <- tibble::tibble(Animal = c(1:n_nr_founder),
                          Sire = rep(NA, n_nr_founder),
                          Dam = rep(NA, n_nr_founder))
tbl_ped
```

The pedigree is also stored in pedigreemm format

```{r}
pem_ped <- pedigreemm::pedigree(sire = tbl_ped$Sire,
                                dam = tbl_ped$Dam,
                                label = tbl_ped$Animal)
pem_ped
```


# Numerator Relationship Matrix
The numerator relationship matrix is computed using the following function ...

The verification with pedigreemm

```{r}
pedigreemm::getA(ped = pem_ped)
```


# Inverse of $A$

```{r}
pedigreemm::getAInv(ped = pem_ped)
```


# Add Animal $3$
Animal $3$ without parents is added to the pedigree

```{r}
(tbl_ped <- dplyr::bind_rows(tbl_ped, tibble::tibble(Animal = 3, Sire = NA, Dam = NA)))
```

The pedigreemm format

```{r}
pem_ped <- pedigreemm::pedigree(sire = tbl_ped$Sire,
                                dam = tbl_ped$Dam,
                                label = tbl_ped$Animal)
```


# Matrices $L^{-1}$ and $D^{-1}$
Matrix $L^{-1}$ is computed as 

$$L^{-1} = I - P$$

The matrix $P$ as $0.5$ on the animal-parent off-diagonal elements. Because in the above pedigree, there are no parent-offspring relations, $P$ is the zero matrix and $L^{-1}$ is the identity matrix. The matrix $D^{-1}$ is the diagonal matrix with elements 

```{r}
1/pedigreemm::Dmat(ped = pem_ped)
```

This leads to the matrix $A^{-1}$ corresponding to the identity matrix. With respect to the rule, for every animal without parents, the element $(D^{-1})_{ii}$ is added to $A^{-1}$. 


# Offspring with One Parent
Adding parents $1$ for animal $3$

```{r}
n_ani_id <- 3
tbl_ped$Sire[n_ani_id] <- 1 
tbl_ped
```

```{r}
pem_ped <- pedigreemm::pedigree(sire = tbl_ped$Sire,
                                dam = tbl_ped$Dam,
                                label = tbl_ped$Animal)
pem_ped
```


# Matrices $L^{-1}$ and $D^{-1}$
Matrix $D^{-1}$ has diagonal elements 

```{r}
mat_D_inv <- diag(1/pedigreemm::Dmat(ped = pem_ped))
mat_D_inv
```

The matrix $(L^{-1})^T$ is given by

```{r}
mat_L_inv <- diag(nrow = n_ani_id)
mat_L_inv[3,1] <- -0.5
t(mat_L_inv)
```

The first two factors multiplied is

```{r}
(mat_LD <- crossprod(mat_L_inv,mat_D_inv))
```

The product with the third factor is 

```{r}
mat_L_inv
```


```{r}
mat_LD %*% mat_L_inv
```

Check

```{r}
pedigreemm::getAInv(ped = pem_ped)
```


# Offspring with Two Parents
Adding parent $2$ for animal $3$

```{r}
n_ani_id <- 3
tbl_ped$Dam[n_ani_id] <- 2
tbl_ped
```

The pedigreemm format

```{r}
pem_ped <- pedigreemm::pedigree(sire = tbl_ped$Sire,
                                dam = tbl_ped$Dam,
                                label = tbl_ped$Animal)
pem_ped
```



# Matrices $L^{-1}$ and $D^{-1}$
Use the following function `get_L_inv()` to obtain the matrix $L^{-1}$

```{r}
get_L_inv <- function(ptbl_ped){
  n_nr_ani <- nrow(ptbl_ped)
  mat_L_inv_result <- diag(nrow = n_nr_ani)
  for (idx in 1:n_nr_ani){
    # sire
    s_idx <- ptbl_ped$Sire[idx]
    if (!is.na(s_idx)) mat_L_inv_result[idx, s_idx] <- -0.5
    # dam
    d_idx <- ptbl_ped$Dam[idx]
    if (!is.na(d_idx)) mat_L_inv_result[idx, d_idx] <- -0.5
  }
  return(mat_L_inv_result)
}
```

The matrix can be obtained as 

```{r}
mat_L_inv <- get_L_inv(ptbl_ped = tbl_ped)
mat_L_inv
```



