---
title: "Design of Problem 1 of Exam - LBGFS 2023"
output: html_notebook
---

```{r setup, include=FALSE}
rm(list = ls())
params <- list(seed = 2023)
```


## Required Package `pedsuite`

```{r}
if (!is.element("pedsuite", installed.packages()))
  install.packages("pedsuite", dependencies = TRUE)
```



## Create Random Pedigree

```{r}
ped_p1c <- pedtools::randomPed(7, founders = 4, seed = params$seed)
plot(ped_p1c)
```

Get labels of animals in the pedigree

```{r}
labels(ped_p1c)
```

Re-label the animals to get order of generations right

```{r}
ped_p1c <- ped_p1c |> pedtools::relabel(new = c("1","2","4","3","5", "7","6"))
plot(ped_p1c)
```




The constructed pedigree

```{r}
library(dplyr)
tbl_ped_p1c <- tibble::as_tibble(ped_p1c)
tbl_ped_p1c <- tbl_ped_p1c %>% 
  select(id, fid, mid) %>%  
  mutate(fid = na_if(fid, "0"),
         mid = na_if(mid, "0")) %>%
  arrange(id)
tbl_ped_p1c
```


Convert the pedigree to numerator relationship matrix

```{r}
pedmm_p1c <- pedigreemm::pedigree(sire = as.integer(tbl_ped_p1c$fid),
                            dam = as.integer(tbl_ped_p1c$mid),
                            label = tbl_ped_p1c$id)
pedmm_p1c
```

```{r}
mat_A <- as.matrix(pedigreemm::getA(ped = pedmm_p1c))
mat_A
```



Write matrix to file

```{r}
s_num_rel_mat_p1 <- here::here("docs", "data", "exam_num_rel_mat_p1.csv")
readr::write_delim(tibble::as_tibble(mat_A), s_num_rel_mat_p1, delim = ",")
```


## Verify

The matrix $P$ is obtained by 

```{r}
nr_animal <- nrow(tbl_ped_p1c)
P = matrix(0, nrow = nr_animal, ncol = nr_animal)
for (i in 1:nr_animal){
  s <- as.integer(tbl_ped_p1c$fid[i])
  d <- as.integer(tbl_ped_p1c$mid[i])
  if (!is.na(s)){
    P[i,s] <- 0.5
  }
  if(!is.na(d)){
    P[i,d] <- 0.5
  }
}
P
```

```{r}
diag_D <- pedigreemm::Dmat(ped = pedmm_p1c)
diag_D
```

The inverse $D^{-1}$

```{r}
mat_D_inv <- diag(1/diag_D)
mat_D_inv
```

The inverse $L^{-1}$ is given by $I-P$

```{r}
mat_L_inv <- diag(nrow = nr_animal) - P
mat_L_inv
```

The matrix $L$ is 

```{r}
mat_L <- solve(mat_L_inv)
mat_L
```

```{r}
mat_L %*% mat_L_inv
```


The cholesky decomposition is 

```{r}
mat_R <- chol(mat_A)
mat_R
```

Check

```{r}
crossprod(mat_R) - mat_A
```

The wrong order of the matrices produces the wrong result, as shown below. 

```{r}
mat_D <- diag(diag_D)
t(mat_L) %*% mat_D %*% mat_L - mat_A
```

The correct order can be verified to produce the correct result

```{r}
mat_D <- diag(diag_D)
mat_L %*% mat_D %*% t(mat_L) - mat_A
```

## Relationship Between Decompositions

From 

$$A = LDL^T$$

and 

$$A = RR^T$$

We get that with the diagonal matrix $S$ such that 

$$R = LS$$

The matrix $S$ is a diagonal matrix with elements $(S)_{ii} = \sqrt{(D)_{ii}}$. From this the matrix $L$ can be computed as 

$$L = RS^{-1}$$


```{r}
mat_S_inv <- diag(1/sqrt(diag_D))
mat_L <- crossprod(mat_R, mat_S_inv)
mat_L
```

To get to the matrix $P$, we have to get the inverse of $L$ 

```{r}
mat_P <- diag(nrow = nr_animal) - solve(mat_L)
round(mat_P, digits = 4)
```




## Take Example Where It Works

```{r}

```

