---
title: "Design of Problem 1 of Exam - LBGFS 2023"
output: html_notebook
---

```{r setup, include=FALSE}
params <- list(seed = 2023)
```


## Required Package `pedsuite`

```{r}
if (!is.element("pedsuite", installed.packages()))
  install.packages("pedsuite", dependencies = TRUE)
```


## Create A complex family tree as a test

```{r}
library(pedtools)

# Top part
x = ancestralPed(g = 2) # bottom person is "7"

# Bottom part
y = halfCousinPed(degree = 1) # top person is "2"
y = swapSex(y, 4)
#> Changing sex of spouses as well: 5

# Merge
z = mergePed(x, y, by = c("7" = "2"), relabel = TRUE)
```



Analyse the single pedigrees

```{r}
# Top part
x = ancestralPed(g = 2) # bottom person is "7"
plot(x)
```

The second part

```{r}
# Bottom part
y = halfCousinPed(degree = 1) # top person is "2"
y = swapSex(y, 4)
#> Changing sex of spouses as well: 5
plot(y)
```

Merge the two pedigrees

```{r}
# Merge
z = mergePed(x, y, by = c("7" = "2"), relabel = TRUE)
plot(z)
```


Example with half-sib pedigrees

```{r}
ped_p1 <- halfSibPed(nch1 = 2, nch2 = 3)
plot(ped_p1)
```

```{r}
ped_p1 <- swapSex(ped_p1,5)
plot(ped_p1)
```

```{r}
ped_p1 <- ped_p1 |> addSon(parents = c(4,5)) |> addDaughter(parents = c(6,5))
plot(ped_p1)
```

More offspring

```{r}
ped_p1 <- ped_p1 |> addDaughter(parents = c(9,10))
plot(ped_p1)
```


## Pedigree As Table To Compute Additive Genetic Relationship

```{r}
tbl_ped_p1 <- tibble::as_tibble(ped_p1)
tbl_ped_p1
```

Filter out founder animals

```{r}
library(dplyr)
tbl_ped_p1_data <- tbl_ped_p1 %>% 
  select(id, fid, mid) %>%
  filter(fid != 0 & mid != 0) %>%
  rename(animal = id, sire = fid, dam = mid)
tbl_ped_p1_data
```


## Write Pedigree Data To File

```{r}
s_ped_p1_data <- here::here("docs", "data", "exam_pedigree_p1.csv")
readr::write_delim(tbl_ped_p1_data, s_ped_p1_data, delim = ",")
```


## Create Random Pedigree

```{r}
ped_p1c <- pedtools::randomPed(7, founders = 4, seed = params$seed)
#ped_p1c <- pedtools::ancestralPed(g=2)
plot(ped_p1c)
```


The constructed pedigree

```{r}
library(dplyr)
tbl_ped_p1c <- tibble::as_tibble(ped_p1c)
tbl_ped_p1c <- tbl_ped_p1c %>% 
  select(id, fid, mid) %>%  
  mutate(fid = na_if(fid, "0"),
         mid = na_if(mid, "0"))
tbl_ped_p1c
```


Convert the pedigree to numerator relationship. matrix

```{r}
ped <- pedigreemm::pedigree(sire = as.integer(tbl_ped_p1c$fid),
                            dam = as.integer(tbl_ped_p1c$mid),
                            label = tbl_ped_p1c$id)
mat_A <- as.matrix(pedigreemm::getA(ped = ped))
mat_A
```

Write matrix to file

```{r}
s_num_rel_mat_p1 <- here::here("docs", "data", "exam_num_rel_mat_p1.csv")
readr::write_delim(tibble::as_tibble(mat_A), s_num_rel_mat_p1, delim = ",")
```


## Verify

The matrix $P$ is obtained by 

```{r}
nr_animal <- nrow(tbl_ped_p1c)
P = matrix(0, nrow = nr_animal, ncol = nr_animal)
for (i in 1:nr_animal){
  s <- as.integer(tbl_ped_p1c$fid[i])
  d <- as.integer(tbl_ped_p1c$mid[i])
  if (!is.na(s)){
    P[i,s] <- 0.5
  }
  if(!is.na(d)){
    P[i,d] <- 0.5
  }
}
P
```

```{r}
diag_D <- pedigreemm::Dmat(ped = ped)
diag_D
```

The inverse $D^{-1}$

```{r}
mat_D_inv <- diag(1/diag_D)
mat_D_inv
```

The inverse $L^{-1}$ is given by $I-P$

```{r}
mat_L_inv <- diag(nrow = nr_animal) - P
mat_L_inv
```

The matrix $L$ is 

```{r}
mat_L <- solve(mat_L_inv)
mat_L
```

The cholesky decomposition is 

```{r}
mat_R <- chol(mat_A)
mat_R
```

Check

```{r}
crossprod(mat_R) - mat_A
```

```{r}
t(mat_L) %*% diag(diag_D) %*% mat_L - mat_A
```



```{r}
t(mat_L) %*% mat_D %*% mat_L - mat_A
```



```{r}
nr_animal
s
```

